Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-06-05 manuscript/manuscript.Rmd
added another figure.

The now extensive corpus of experimental archaeology have principally focused on the generation of knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999]. There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.

1)  True controlled experiments are rare. Randomized controlled experiments *sensu stricto* are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past are derived from datasets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023]. It has also been a debatable issue in philosophy of science whether Randomized Controled Trial (RCT) represents the golden standard of knowledge [@cartwright2007].
2)  Trade-off between causality and generalizability in experimental design. For example, in a series of recently published experimental studies aiming at understanding the role of language in the transmission of lithic technologies, standardized procedures like video teaching or using bricks as raw materials are very often.

In the past decades, actualistic experiments becomes more common [@liu2022]. Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [@stout2002; @arthur2018].

The process level can be dismantled into two parts: 1)the cognitive [@stout2015] and physical demands [@key2018] of certain tool-making behavior; 2) the ethological analysis.

@@

Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conduct systematical ethogram analysis. With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years. Here we use action grammar developed by [@stout2021] as an example. Other coding scheme also exist such as [@mahaney2014], or rotation analysis [@muller2023], or [@cueva-temprana2019].

>>>>

The now extensive corpus of experimental archaeology have witnessed the growing principally focused on the generation of knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999]. There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.

Controlled experiments without randomization are not enough to infer causal mechanism, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012]. Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past are derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through expeirments featuring small-sized convenience sample. It has also been a debatable issue whether Randomized Controled Trial (RCT) represents the golden standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].

Trade-off between causality and generalizability in experimental design. For example, in a series of recently published experimental studies aiming at understanding the role of language in the transmission of lithic technologies, standardized procedures like video teaching or using bricks as raw materials are very often.

In the past decades, actualistic experiments becomes more common [@liu2022]. Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [@stout2002; @arthur2018]. Statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023].

@@

Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conduct systematical ethogram analysis. With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years (**Figure** \@ref(fig:ethogram)). Here we use action grammar developed by [@stout2021] as an example. Other coding scheme also exist such as [@mahaney2014], or rotation analysis [@muller2023], or [@cueva-temprana2019].

```{r ethogram, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="An example of coding Bruce Bradley's handaxe knapping session using the action grammer and BORIS software."}
knitr::include_graphics(here::here("figure/Fig2.jpg"))
```
